Review: Adaptive Introgression Expanded the Genetic Base of Crops during post-Domestication Spread

### 2017-07-07 paper.tex
Update paper.tex
edits to rice case study

\item{Sunflower: }

The common sunflower (\emph{Helianthus annuus}) shows evidence of domestication in the eastern United States \cite{harter2004origin, wills2006chloroplast}\, with potential for a second domestication center in Mexico \cite{lentz2008sunflower}.
The pre-Columbian \emph{H. annuus} distribution of cultivated sunflower spanned much of the Great Plains, from what is now north-central Texas to Montana and North Dakota (see figure 1 of \cite{whitney2010adaptive}).
Domesticated sunflower has long lived in sympatry with wild relatives such as \emph{H. petiolaris} and \emph{H. bolanderi} and forms stable hybrid populations with these taxa \cite{schwarzbach2002likely, rieseberg1988molecular, welch2002patterns}.
Wild sunflowers are known to be locally-adapted, and weedy hybrid populations often share these adaptations \cite{kane2008genetics}.
However, the most striking example of adaptive introgression within \emph{Helianthus} is that of the cucumberleaf sunflower, \emph{H. debilis} ssp. \emph{cucumerifolius}.
\mbh{I believe \emph{H. annuus} ssp. \emph{texanus} is a wild sunflower, right?  If so, this is more an example of hybrid speciation rather then adaptive wild-to-crop gene flow}
Cucumberleaf sunflower is endemic to south-central Texas, and exhibits several adaptations to the region.
Introgressive hybridization imparted locally-adapted alleles from \emph{H. debilis} to \emph{H. annuus} via introgressive hybridization \cite{heiser1951hybridization}. 
These introgressed hybrids formed a new lineage of sunflower (\emph{H. annuus} ssp. \emph{texanus}, \emph{H. a. texanus} hereafter) which displays \emph{H. debilis}-like traits adaptive to south-central Texas climate and ecology.
These adaptive \emph{debilis}-like traits include resistance to herbivorous pests and an increased branching plant architecture, as well as higher overall fitness than \emph{H. annuus} (as measured by higher seed production \cite{whitney2006adaptive}).
Although H. annuus and \emph{H. a. texanus} are interfertile, \emph{H. a. texanus} displays persistent phenotypic differences from \emph{H. annuus} \cite{rieseberg2007hybridization}.

The genome of the common sunflower has been greatly influenced by introgression from wild relatives, due to both natural outcrossing events and concerted breeding efforts in crop improvement.
\emph{Helianthus} has several genes for downy mildew resistance, and each imparts resistance to one or more races of \emph{Plasmopara halstedii}, one of the most agronomically important diseases in sunflower cultivation \cite{cohen1973factors}.
Some of these downy mildew resistence genes were found in wild relatives (including \emph{H. argophyllus}, \emph{H. tuberosus}, and \emph{H. praecox}) and have been successfully bred into modern \emph{H. annuus} \cite{miller1991inheritance}.
PlArg, an allele found in wild silverleaf sunflowers (\emph{H. argophyllus}, inbred line Arg1575-2), confers resistance to all known (20 or more) races of downey mildew \cite{dussle2004pl}\, while others (Pl1-Pl11) are effective for one or more types \cite{rahim2002inheritance}.
Silverleaf sunflower has also been the focus of drought resistance breeding efforts \cite{saucă2010introgression}\ and \emph{Phomopsis} resistance breeding efforts \cite{besnard1997specifying}.
\emph{H. annuus} shows signs of persistent introgressive hybridization with \emph{H. petiolaris} with evidence of positive selection driving some of the genetic differentiation between the two species \cite{yatabe2007rampant}.

Recent investigations into the history of \emph{Helianthus} introgression have implemented genomic methods.
\cite{Baute2015} analyzed transcriptome sequence variation on cultivated and wild \emph{H. annuus}, \emph{H. petiolaris}, and \emph{H. argophyllus}.
Using STRUCTURE, these authors found that introgressions from wild relatives exist on every chromosome in at least one modern line, covering over 10\% of the genome.
Of particular note is the modern line RHA 274, a modern line which was bred with \emph{H. a. texanus} in the 1970s to restore a branching plant body architecture, which allows the plant to produce pollen for a longer period of time, increasing seed production.
RHA 274 has several large introgression from \emph{H. a. texanus}, including one at the site of HaGNAT, the domestication gene associated with branching.
These introgressed regions are not found in the non-branching lines Sunrise and VNIIMK8931, further suggesting that the \emph{H. a. texanus} introgressed regions are causative.

Notes for review:

http://biology.unm.edu/Whitney/Whitney%20Reprints/Whitney%20et%20al.%202006%20Am%20Nat.pdf

http://www.jstor.org/stable/pdf/2409227.pdf?refreqid=excelsior:a686b7f06f9daa2b89bd5a654dade3b4

@@

http://onlinelibrary.wiley.com/doi/10.1111/j.1469-8137.2010.03234.x/epdf

https://biology.unm.edu/Whitney/Whitney%20Reprints/Scascitelli%20et%20al%202010.pdf

Genome scan of hybridizing sunflowers from Texas
(Helianthus annuus and H. debilis) reveals asymmetric
patterns of introgression and small islands of genomic
differentiation:

In contrast, coalescent analyses of long-term migration
rates imply that interspecific migration has been
more important than mutation in providing new genetic
variation in ssp. texanus populations. Immigration from
H. debilis cucumerifolius into ssp. texanus (M = 4.05;
Nem = 1.74) is higher than median estimates of intraspecific
migration rates in plants (Morjan and Rieseberg 2004) 
and only slightly lower than immigration from
ssp. annuus into ssp. texanus (M = 5.00). Immigration
rates from ssp. texanus into H. debilis cucumerifolius and
ssp. annuus are still higher (M = 6.30 and 7.29, respectively),
implying that the putative hybrid lineage serves
as a bridge for migration between the parental species.

One puzzle is
that the estimated immigration rate between the allopatric
taxa (H. debilis cucumerifolius and ssp. annuus) is
approximately comparable with that between the sympatric
taxa (H. debilis cucumerifolius and ssp. texanus ssp.
texanus). This observation may point to the effectiveness
of ssp. texanus as a bridge for the transfer of alleles
between H. debilis and ssp. annuus. An alternative explanation
is that the BAYESASS estimates are unreliable
because homoplasy of microsatellite alleles is high in
taxa with large-effective population sizes (Estoup et al.
2002) such as the annual sunflowers studied here.

Despite the apparent discrepancy between the results
from STRUCTURE and BAYEASS, several conclusions
can be made. Analyses with STRUCTURE support the
distinctness of ssp. texanus, as well as its clear inclusion
within H. annuus, as initially postulated by Heiser (1951).
Likewise, the STRUCTURE, MIGRATE-N and BAYESASS
analyses all confirm previous report of introgression
between H. annuus and H. debilis (Rieseberg et al.
1990, 2007) and suggest that the introgression is genomewide
when it occurs. Contrary to the Heiser’s scenario of
a recent Holocene origin of ssp. texanus, however, our
results are more consistent with a much longer history of
contact between H. annuus and H. debilis. Otherwise, it is
difficult to account for the presence of significant longterm
migration between currently allopatric populations
of ssp. annuus and H. debilis cucumerifolius. This revised
scenario makes sense in the light of numerous glacialinterglacial
cycles over the past million years (Hewitt
2000), which probably resulted in intermittent contact
between H. annuus and H. debilis, with the last contact
likely occurring during the Wisconsin glaciation,
18 000 BP. The current contact between the species differs
from the previous ones in that it appears to have been
human-aided. Possibly, some of the molecular evidence
of introgression, particularly into allopatric populations,
stems from past periods of contact.

One of the four outlier loci, HT0414, is associated
with a heat-shock protein and was previously shown to
have been the target of selection in H. annuus ssp. annuus
salt marsh populations from the state of Utah
(Kane and Rieseberg 2007), as well as in weedy sunflower populations
from several different locations across the USA
(Kane and Rieseberg 2008). In this study, populations of
ssp. annuus from Kansas, Nebraska and Oklahoma
monomorphic for a single allele, whereas populations
of H. debilis and other populations of ssp. annuus are
considerably more polymorphic. These results imply
that HT0414 may be involved in adaptation to a range
of different habitats or to conditions shared by several
different habitats (Kane and Rieseberg 2008).

@@

The story of Asian rice (\emph{Oryza sativa}) domestication is still debated.
Interspecific hybridization between wild and domesticated rices both ancient and recent has produced a phylogeny that defies clear and simple models.
On one hand, some genetic and archaeobotanical evidence point toward independent domestications of the two prominent varietal groups \emph{japonica} and \emph{indica} from the wild species \emph{Oryza rufipogon} (\emph{rufipogon} hereafter) in the Yangtze Basin and the Indian Ganges plain, respectively \cite{fuller2010consilience}.
Other studies support a single center of domestication, arguing that present-day patterns of rice variation could be explained by a single domestication event.
A recent investigation utilizing SNP data from wild and domesticated rice accesssions pointed to a single domestication occurring 8,200-13,500 BP in the Yangtze Basin in China from rufipogon, with later divergence of \emph{indica} and \emph{japonica} \cite{molina2011molecular}.
Huang and colleagues \cite{Huang2012} developed a genetic map of rice variation, which they used to measure genetic distance between wild and domesticated rices at and around domesication loci for various geographical locations, finding that \emph{japonica} was likely domesticated in the Pearl River area of Guangxi province, China (just south of the Yangtze Basin), and that \emph{indica} was likely the result of hybridization between \emph{japonica} and local \emph{rufipogon} populations in Southern and South-eastern Asia. %Also, Huang includes the following quote: "We noticed that many loci with strong signals of selection were nearly identical in both indica and japonica where FST between indica and japonica was extremely low, indicating that introduction of traits during domestication has in many cases involved introgression events."

@@

\subsection*{What candidate genes were targeted by selection during domestication?}
Loci targeted by selection during domestication can be identified through so-called ``bottom-up'' approaches based on population genetic signatures \cite{Ross-Ibarra2007}.
Ideally, candidate loci will be identified by first constructing a demographic model representing the history of the domesticate.
In this approach, diversity data from neutral loci are fit to potential models of a crop's demography and then statistical tests of selection are used to identify candidate domestication genes under the most likely model.
Due to the difficulty of this approach and the uncertainty associated with any given demographic model, many studies identify domestication loci using a strict outlier approach in which loci showing the greatest reduction in nucleotide diversity or the highest allele frequency differentiation in the domesticate relative to the wild progenitor are identified as candidates.
Introgression during crop expansion may influence candidate gene detection using both demographic-modeling and strict-outlier approaches.
For example, \emph{mexicana} introgression into maize described above accounts for approximately 20\% of the genome of maize in the highlands of Mexico \cite{vanHeerwaarden2011}.
Takuno and co-authors \cite{Takuno2015} have shown that a demographic model incorporating this introgression is a significantly better fit to empirical data than a model lacking introgression.
Failure to account for introgression in maize would therefore compromise candidate detection, particularly if a study contained maize samples from the Mexican highlands.
Likewise, introgression that increased nucleotide diversity in the domesticate or decreased differentiation at domestication loci would confound a strict outlier approach.
However, previous work, also in maize, has shown that known domestication loci are particularly resistant to introgression \cite{hufford2013genomic}, likely due to ongoing selection favoring the domesticated phenotype.
\gmj{This is true in most cases I think, although I remember reading that an opposite pattern was found in sunflowers.  I'll go find that paper again and see if it warrants a mention here.}

>>>>

%\item{Sunflower: }

%The common sunflower (\emph{Helianthus annuus}) shows evidence of domestication in the eastern United States \cite{harter2004origin, wills2006chloroplast}\, with potential for a second domestication center in Mexico \cite{lentz2008sunflower}.
%The pre-Columbian \emph{H. annuus} distribution of cultivated sunflower spanned much of the Great Plains, from what is now north-central Texas to Montana and North Dakota (see figure 1 of \cite{whitney2010adaptive}).
%Domesticated sunflower has long lived in sympatry with wild relatives such as \emph{H. petiolaris} and \emph{H. bolanderi} and forms stable hybrid populations with these taxa \cite{schwarzbach2002likely, rieseberg1988molecular, welch2002patterns}.
%Wild sunflowers are known to be locally-adapted, and weedy hybrid populations often share these adaptations \cite{kane2008genetics}.
%However, the most striking example of adaptive introgression within \emph{Helianthus} is that of the cucumberleaf sunflower, \emph{H. debilis} ssp. \emph{cucumerifolius}.
%\mbh{I believe \emph{H. annuus} ssp. \emph{texanus} is a wild sunflower, right?  If so, this is more an example of hybrid speciation rather then adaptive wild-to-crop gene flow}
%Cucumberleaf sunflower is endemic to south-central Texas, and exhibits several adaptations to the region.
%Introgressive hybridization imparted locally-adapted alleles from \emph{H. debilis} to \emph{H. annuus} via introgressive hybridization \cite{heiser1951hybridization}. 
%These introgressed hybrids formed a new lineage of sunflower (\emph{H. annuus} ssp. \emph{texanus}, \emph{H. a. texanus} hereafter) which displays \emph{H. debilis}-like traits adaptive to south-central Texas climate and ecology.
%These adaptive \emph{debilis}-like traits include resistance to herbivorous pests and an increased branching plant architecture, as well as higher overall fitness than \emph{H. annuus} (as measured by higher seed production \cite{whitney2006adaptive}).
%Although H. annuus and \emph{H. a. texanus} are interfertile, \emph{H. a. texanus} displays persistent phenotypic differences from \emph{H. annuus} \cite{rieseberg2007hybridization}.

%The genome of the common sunflower has been greatly influenced by introgression from wild relatives, due to both natural outcrossing events and concerted breeding efforts in crop improvement.
%\emph{Helianthus} has several genes for downy mildew resistance, and each imparts resistance to one or more races of \emph{Plasmopara halstedii}, one of the most agronomically important diseases in sunflower cultivation \cite{cohen1973factors}.
%Some of these downy mildew resistence genes were found in wild relatives (including \emph{H. argophyllus}, \emph{H. tuberosus}, and \emph{H. praecox}) and have been successfully bred into modern \emph{H. annuus} \cite{miller1991inheritance}.
%PlArg, an allele found in wild silverleaf sunflowers (\emph{H. argophyllus}, inbred line Arg1575-2), confers resistance to all known (20 or more) races of downey mildew \cite{dussle2004pl}\, while others (Pl1-Pl11) are effective for one or more types \cite{rahim2002inheritance}.
%Silverleaf sunflower has also been the focus of drought resistance breeding efforts \cite{saucă2010introgression}\ and \emph{Phomopsis} resistance breeding efforts \cite{besnard1997specifying}.
%\emph{H. annuus} shows signs of persistent introgressive hybridization with \emph{H. petiolaris} with evidence of positive selection driving some of the genetic differentiation between the two species \cite{yatabe2007rampant}.

%Recent investigations into the history of \emph{Helianthus} introgression have implemented genomic methods.
%\cite{Baute2015} analyzed transcriptome sequence variation on cultivated and wild \emph{H. annuus}, \emph{H. petiolaris}, and \emph{H. argophyllus}.
%Using STRUCTURE, these authors found that introgressions from wild relatives exist on every chromosome in at least one modern line, covering over 10\% of the genome.
%Of particular note is the modern line RHA 274, a modern line which was bred with \emph{H. a. texanus} in the 1970s to restore a branching plant body architecture, which allows the plant to produce pollen for a longer period of time, increasing seed production.
%RHA 274 has several large introgression from \emph{H. a. texanus}, including one at the site of HaGNAT, the domestication gene associated with branching.
%These introgressed regions are not found in the non-branching lines Sunrise and VNIIMK8931, further suggesting that the \emph{H. a. texanus} introgressed regions are causative.

%Notes for review:

%http://biology.unm.edu/Whitney/Whitney%20Reprints/Whitney%20et%20al.%202006%20Am%20Nat.pdf

%http://www.jstor.org/stable/pdf/2409227.pdf?refreqid=excelsior:a686b7f06f9daa2b89bd5a654dade3b4

@@

%http://onlinelibrary.wiley.com/doi/10.1111/j.1469-8137.2010.03234.x/epdf

%https://biology.unm.edu/Whitney/Whitney%20Reprints/Scascitelli%20et%20al%202010.pdf

%Genome scan of hybridizing sunflowers from Texas
%(Helianthus annuus and H. debilis) reveals asymmetric
%patterns of introgression and small islands of genomic
%differentiation:

%In contrast, coalescent analyses of long-term migration
%rates imply that interspecific migration has been
%more important than mutation in providing new genetic
%variation in ssp. texanus populations. Immigration from
%H. debilis cucumerifolius into ssp. texanus (M = 4.05;
%Nem = 1.74) is higher than median estimates of intraspecific
%migration rates in plants (Morjan and Rieseberg 2004) 
%and only slightly lower than immigration from
%ssp. annuus into ssp. texanus (M = 5.00). Immigration
%rates from ssp. texanus into H. debilis cucumerifolius and
%ssp. annuus are still higher (M = 6.30 and 7.29, respectively),
%implying that the putative hybrid lineage serves
%as a bridge for migration between the parental species.

%One puzzle is
%that the estimated immigration rate between the allopatric
%taxa (H. debilis cucumerifolius and ssp. annuus) is
%approximately comparable with that between the sympatric
%taxa (H. debilis cucumerifolius and ssp. texanus ssp.
%texanus). This observation may point to the effectiveness
%of ssp. texanus as a bridge for the transfer of alleles
%between H. debilis and ssp. annuus. An alternative explanation
%is that the BAYESASS estimates are unreliable
%because homoplasy of microsatellite alleles is high in
%taxa with large-effective population sizes (Estoup et al.
%2002) such as the annual sunflowers studied here.

%Despite the apparent discrepancy between the results
%from STRUCTURE and BAYEASS, several conclusions
%can be made. Analyses with STRUCTURE support the
%distinctness of ssp. texanus, as well as its clear inclusion
%within H. annuus, as initially postulated by Heiser (1951).
%Likewise, the STRUCTURE, MIGRATE-N and BAYESASS
%analyses all confirm previous report of introgression
%between H. annuus and H. debilis (Rieseberg et al.
%1990, 2007) and suggest that the introgression is genomewide
%when it occurs. Contrary to the Heiser’s scenario of
%a recent Holocene origin of ssp. texanus, however, our
%results are more consistent with a much longer history of
%contact between H. annuus and H. debilis. Otherwise, it is
%difficult to account for the presence of significant longterm
%migration between currently allopatric populations
%of ssp. annuus and H. debilis cucumerifolius. This revised
%scenario makes sense in the light of numerous glacialinterglacial
%cycles over the past million years (Hewitt
%2000), which probably resulted in intermittent contact
%between H. annuus and H. debilis, with the last contact
%likely occurring during the Wisconsin glaciation,
%18 000 BP. The current contact between the species differs
%from the previous ones in that it appears to have been
%human-aided. Possibly, some of the molecular evidence
%of introgression, particularly into allopatric populations,
%stems from past periods of contact.

%One of the four outlier loci, HT0414, is associated
%with a heat-shock protein and was previously shown to
%have been the target of selection in H. annuus ssp. annuus
%salt marsh populations from the state of Utah
%(Kane and Rieseberg 2007), as well as in weedy sunflower populations
%from several different locations across the USA
%(Kane and Rieseberg 2008). In this study, populations of
%ssp. annuus from Kansas, Nebraska and Oklahoma
%monomorphic for a single allele, whereas populations
%of H. debilis and other populations of ssp. annuus are
%considerably more polymorphic. These results imply
%that HT0414 may be involved in adaptation to a range
%of different habitats or to conditions shared by several
%different habitats (Kane and Rieseberg 2008).

@@

The story of Asian rice (\emph{Oryza sativa}) domestication is still debated, with hybridization between wild and domesticated rice contributing to the complexity of this crop's history.
On one hand, some genetic and archaeobotanical evidence point toward independent domestications of the two prominent varietal groups \emph{japonica} and \emph{indica} from the wild species \emph{Oryza rufipogon} (\emph{rufipogon} hereafter) in the Yangtze Basin and the Indian Ganges plain, respectively \cite{fuller2010consilience}.
Other studies support a single domestication occurring 8,200-13,500 BP in the Yangtze Basin in China from \emph{rufipogon}, with later divergence of \emph{indica} and \emph{japonica} \cite{molina2011molecular}.
Huang and colleagues \cite{Huang2012} developed a genetic map of rice variation, which they used to measure genetic distance between wild and domesticated rices at and around domesication loci for various geographical locations, finding that \emph{japonica} was likely domesticated in the Pearl River area of Guangxi province, China (just south of the Yangtze Basin), and that \emph{indica} was likely the result of hybridization between \emph{japonica} and local \emph{rufipogon} populations in Southern and South-eastern Asia. %Also, Huang includes the following quote: "We noticed that many loci with strong signals of selection were nearly identical in both indica and japonica where FST between indica and japonica was extremely low, indicating that introduction of traits during domestication has in many cases involved introgression events."

@@

\subsection*{What candidate genes were targeted by selection during domestication?}
Loci targeted by selection during domestication can be identified through so-called ``bottom-up'' approaches based on population genetic signatures \cite{Ross-Ibarra2007}.
Ideally, candidate loci will be identified by first constructing a demographic model representing the history of the domesticate.
In this approach, diversity data from neutral loci are fit to potential models of a crop's demography and then statistical tests of selection are used to identify candidate domestication genes under the most likely model.
Due to the difficulty of this approach and the uncertainty associated with any given demography, many studies identify domestication loci using a strict outlier approach in which loci showing the greatest reduction in nucleotide diversity or the highest allele frequency differentiation in the domesticate relative to the wild progenitor are identified as candidates.
Introgression during crop expansion may influence candidate gene detection using both demographic-modeling and strict-outlier approaches.
For example, \emph{mexicana} introgression into maize described above accounts for approximately 20\% of the genome of maize in the highlands of Mexico \cite{vanHeerwaarden2011}.
Takuno and co-authors \cite{Takuno2015} have shown that a demographic model incorporating this introgression is a significantly better fit to empirical data than a model lacking introgression.
Failure to account for introgression in maize would therefore compromise domestication candidate detection, particularly if a study contained maize samples from the Mexican highlands.
Likewise, introgression that increased nucleotide diversity in the domesticate or decreased differentiation at domestication loci would confound a strict outlier approach.
However, previous work, also in maize, has shown that known domestication loci are particularly resistant to introgression \cite{hufford2013genomic}, likely due to ongoing selection favoring the domesticated phenotype.
\gmj{This is true in most cases I think, although I remember reading that an opposite pattern was found in sunflowers.  I'll go find that paper again and see if it warrants a mention here.}
